---
title: "Working with Genomic Data"
output: 
  rmarkdown::github_document:
  toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

A tutorial for analysing and processing genomic data using simple-to-use R tools

# Table of contents
- [Getting started](#Getting-started)
  * [Installing R](#Installing-R)
  * [Installing packages for this tutorial](#Installing-packages-for-this-tutorial)
- [Importing data into R](#Importing-data-into-R)
- [Transforming dataframes](#Transforming-dataframes)
  * [The pipe (%>%)](#The-pipe-(%>%))
  * [Filter data](#Filter-data)
  * [Transform data](#Transform-data)
  * [Reformating columns](##Reformating-columns)
  * [Spread/gather](#Spreadgather)

# Getting started
## Installing R
To begin this tutorial, you need to have R installed on your computer. You may download the latest R version so from [CRAN](https://cloud.r-project.org/).

Optionally, you may enhance your R experience by utilizing Rstudio, an easy to use Integrated Development Environment (IDE) while running R code. Rstudio can be downloaded from [here](https://rstudio.com/products/rstudio/).


## Installing packages for this tutorial
R contain a plethora of effecient and simple tools for data analysis. To install all the packages covered in this tutorial, open R in your terminal or from Rstudio and run the following:

```{r install, eval=FALSE}
install.packages("devtools")
devtools::install_github("fursham-h/wwgd")
```

## Importing data into R






## Transforming dataframes
### The pipe (%>%)
### Filter data
### Transform data
### Reformating columns
### Spread/gather
