---
title: "Working with Data"
output: 
  rmarkdown::github_document:
  toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r silentload, include=FALSE}
#Silent load all dependencies for vignette
library(vroom)
library(readxl)
library(tidyverse)
```
A tutorial for analysing and processing data using simple-to-use R tools

# Table of contents
- [Getting started](#Getting-started)
  * [Installing R](#Installing-R)
  * [Installing packages for this tutorial](#Installing-packages-for-this-tutorial)
- [Importing data into R](#Importing-data-into-R)
  * [Retrieving path to your data file](#Retrieving-path-to-your-data-file)
  * [Importing delimited text files](#Importing-delimited-text-files
  * [Importing excel files](#Importing-excel-files)
- [Transforming dataframes](#Transforming-dataframes)
  * [The pipe (%>%)](#The-pipe-(%>%))
  * [Filter data](#Filter-data)
  * [Transmute data](#Transmute-data)
  * [Reformating columns](##Reformating-columns)
  * [Spread/gather](#Spreadgather)

## Getting started
### Installing R
To begin this tutorial, you need to have R installed on your computer. You may download the latest R version so from [CRAN](https://cloud.r-project.org/).

Optionally, you may enhance your R experience by utilizing Rstudio, an easy to use software to run and write R codes. Rstudio can be downloaded from [here](https://rstudio.com/products/rstudio/).


### Installing packages for this tutorial
R contain a plethora of effecient and simple tools for data analysis. To install all the packages covered in this tutorial, open R in your terminal or from Rstudio and run the following:

```{r install, eval=FALSE}
install.packages("devtools")
devtools::install_github("fursham-h/wwgd")
```

## Getting data into R
Most biology data are structured as tables. It contain one or more column variables and an infinite number of rows listing the values of each variable. A data table can be saved as one of these file types; 

* generic text file: *.txt*
* tab-separated values file: *.tsv*
* comma-separated values file: *.csv*
* excel file: *.xls or .xlsx*. 

To work with these data files in R, we need to import it and convert into an R-friendly dataframe. Here we will describe how to get your data into R.

### Retrieving path to your data file
Before importing your data into R, you need to know the "file path" of your file. You can think of file path as an address in the file system, pointing to where your it is located on your computer system. For example, a file named "testfile.txt" that is located on your user desktop would have a file path that would look something like this:
```{r path, echo=FALSE, comment=''}
cat("/Users/USERNAME/Desktop/testfile.txt")
```

Retrieving your file path differs between operating systems. You may refer to these tips on how to copy file paths for [Mac](http://osxdaily.com/2015/11/05/copy-file-path-name-text-mac-os-x-finder/) and [Windows](https://www.pcworld.com/article/251406/windows-tips-copy-a-file-path-show-or-hide-extensions.html). Once your file path have been copied, you may paste it into a variable in R as such:
```{r savepath, eval=FALSE}
file_path <- "/PATH/TO/YOUR/FILE.txt"
```


A simpler and interactive way of retrieving a file path is possible using R's `file.choose` function. Try:
```{r retrievepathinteractive,eval=F}
file_path <- file.choose()
```


For this part of the tutorial, we will retrieve the path of a sample *.txt* to be imported into R:
```{r retrievepath}
sample_file_path <- system.file("extdata", "expression_1.txt", package = "wwgd")

# inspect path
sample_file_path
```

### Importing delimited text files
To import *.txt*, *.csv* or *.tsv* files into R as a dataframe, I recommend using the `vroom` data import tool from [*vroom*](https://github.com/r-lib/vroom) package. This highly popular tool has plenty of handy features, including guessing of your file delimeter, guessing of your column type, importing of multiple files and most importantly, it is FAST. Simply run:
```{r importxt, message=FALSE}
library(vroom)
sample_file <- vroom(sample_file_path)
```

We can quickly inspect the dataframe using built-in R functions:
```{r inspectfile}
# get overview of the variables in the dataframe
summary(sample_file)
```

As seen above, our sample *expression_1.txt* file contain 2 named columns; "transcript_id" which contain character names of transcripts and "TpM" containing expression values. By default, `vroom` will assign the first row of the input table as column names and the remaining rows as values. If your data do not contain column names, you may prevent `vroom` from misreading your data by setting the **col_names** argument to FALSE as such:
```{r shownoheader,message=FALSE}
vroom(sample_file_path, col_names = FALSE)
```

For more information on how to use *vroom*, refer to [*vroom*](https://vroom.r-lib.org/articles/vroom.html)'s getting started page.

### Importing excel files

To import excel files, I recommend using the `read_excel` function from the [*readxl*](https://readxl.tidyverse.org/) package. Here's how to import an excel version of *expression_1.txt* file into R:
```{r importexcel, message=FALSE}
library(readxl)

# get filepath
sample_file_path <- sample_file_path <- system.file("extdata", "expression_1.xlsx", package = "wwgd")

# import
sample_file <- read_excel(sample_file_path)

# inspect file summary
summary(sample_file)
```

By default, `read_excel` will import the first sheet of your excel file. It is possible to import other sheets by providing additional arguments to this function. Check out *readxl*'s overview page [here](https://readxl.tidyverse.org/).

## Accessing data in dataframes
Now that your data is in R, we can start to inspect the dataframe and extract information from it. We will be using the **sample_file** dataframe that was loaded [above](#Importing-excel-files).

### View contents of dataframe
### Check column/row names and lengths
### Subsetting dataframe


## Transforming dataframes
After importing of data table into R's dataframe, you would want to filter/reshape/transmute your dataframe to better understand your data. Here we will be using simple tools from the popular *tidyverse* package to transform your data

```{r loadtidy, eval=FALSE}
#load tidyverse
library(tidyverse)
```

### The pipe (%>%)

https://www.datacamp.com/community/tutorials/pipe-r-tutorial

### Filter data
### Transmute data
### Reformating columns
### Spread/gather
